---
title: "Part 2 - Normalize the Database"
author: "JC Harrop"
date: "29/01/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Now that we have created a database and loaded something into it from several R dataframes.  These dataframes
had been read from CSV format files which were exported from Excel spreadsheets where field notes and 
instrument readings were being maintained.  Note that within the `tidyverse` environment we could have used 
`dplyr` to read directly from an Excel spreadsheet file, but the example followed CSV format files since 
that is a more widely used data transfer format.

## What is Normalization?

Relational databases are designed to operate effeciently with structured data where various rules governing
data integrity are enforced.

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
